---
title: "260_Proj2"
author: "Yi-Syuan Yen"
date: "2025-02-01"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(lme4)
library(gt)
library(ordinal)
```

## Data preprocessing

We have missing completely at random. Thus, we delete rows with missing entries or plot numbers.

We combine the datasets of disease scores and strawberries traits.

There are 3 datasets: `df.merge.bot`, `df.merge.col`, `df.merge.NPT`, for the 3 diseases.

```{r}
brix = read.csv("/Users/ysyen/Desktop/260/proj2/2024_Consolidated_BrixAcid.csv")
col = read.csv("/Users/ysyen/Desktop/260/proj2/Clean_Col_Ratings.csv")
bot = read.csv("/Users/ysyen/Desktop/260/proj2/Clean_Bot_Ratings.csv")
NPT = read.csv("/Users/ysyen/Desktop/260/proj2/Clean_NPT_Ratings.csv")
f = read.csv("/Users/ysyen/Desktop/260/proj2/Firmness.csv")

col = col %>% select(Score, HarvestDate, Entry, Block, HarvestPeriod)
bot = bot %>% select(Score, HarvestDate, Entry, Block, HarvestPeriod)
NPT = NPT %>% select(Score, HarvestDate, Entry, Block, HarvestPeriod)

col = col %>% filter(is.na(Entry)==F)
bot = bot %>% filter(is.na(Entry)==F)
NPT = NPT %>% filter(is.na(Entry)==F)
# Entry denotes different cultivars
```

```{r}
# Missing rows
# print("bot")
# rownames(bot[is.na(bot$Entry),])
# 
# print("NPT")
# rownames(NPT[is.na(NPT$Entry),])
```


```{r}
brix_sim = brix[, c("Harvest_Date", "Plot", "Block","Entry","Brix", "CorrectedTitro")]
f_sim = f[, c("HarvestPeriod", "Plot", "Block","Entry","Firmness.kg")]
colnames(brix_sim)[1] = "HarvestDate"

f_mean<- aggregate(x= f_sim$Firmness.kg,
                      # Specify group indicator
                      by = list(f_sim$HarvestPeriod,f_sim$Plot,f_sim$Block,f_sim$Entry),      
                      # Specify function (i.e. mean)
                      FUN = mean)
colnames(f_mean) = colnames(f_sim)

df.merge.col = merge(col, brix_sim, by = c("Block","Entry","HarvestDate"), all.x = TRUE)
df.merge.bot = merge(bot, brix_sim, by = c("Block","Entry","HarvestDate"), all.x = TRUE)
df.merge.NPT = merge(NPT, brix_sim, by = c("Block","Entry","HarvestDate"), all.x = TRUE)

df.merge.col = df.merge.col %>% filter(is.na(Plot)==F)
df.merge.bot = df.merge.bot %>% filter(is.na(Plot)==F)
df.merge.NPT = df.merge.NPT %>% filter(is.na(Plot)==F)

df.merge.col = merge(df.merge.col, f_mean, 
                     by = c("Block","Entry","HarvestPeriod","Plot"), all.x = TRUE)
df.merge.bot = merge(df.merge.bot, f_mean, 
                     by = c("Block","Entry","HarvestPeriod","Plot"), all.x = TRUE)
df.merge.NPT = merge(df.merge.NPT, f_mean, 
                     by = c("Block","Entry","HarvestPeriod","Plot"), all.x = TRUE)

df.merge.col$Block = factor(df.merge.col$Block)
df.merge.col$Entry = factor(df.merge.col$Entry)
df.merge.col$HarvestPeriod = factor(df.merge.col$HarvestPeriod)

df.merge.bot$Block = factor(df.merge.bot$Block)
df.merge.bot$Entry = factor(df.merge.bot$Entry)
df.merge.bot$HarvestPeriod = factor(df.merge.bot$HarvestPeriod)

df.merge.NPT$Block = factor(df.merge.NPT$Block)
df.merge.NPT$Entry = factor(df.merge.NPT$Entry)
df.merge.NPT$HarvestPeriod = factor(df.merge.NPT$HarvestPeriod)

```

```{r}
rm(col,NPT,brix,bot)
```

## Data cleaning

```{r}
c.df = df.merge.bot %>% group_by(HarvestPeriod,Block,Entry) %>%tally() 
# table(c.df$Entry)[table(c.df$Entry)==6]
df.merge.bot2 = df.merge.bot[df.merge.bot$Entry %in% names(table(c.df$Entry)[table(c.df$Entry)==6]),]
```


## EDA

```{r}
# Score distribution
ggplot(df.merge.bot, aes(x=as.numeric(Score))) + 
  geom_bar()+
  theme_bw()+
  xlab("Score")+
  ggtitle("Barplot of Score(BOT)")

ggplot(df.merge.col, aes(x=as.factor(Score))) + 
  geom_bar()+
  theme_bw()+
  xlab("Score")+
  ggtitle("Barplot of Score(COL)")

ggplot(df.merge.NPT, aes(x=as.factor(Score))) + 
  geom_bar()+
  theme_bw()+
  xlab("Score")+
  ggtitle("Barplot of Score(NPT)")

```


```{r}
# number of observation per cultivar/block/harvest/Entry
obs = data.frame(table(df.merge.bot$Entry))
hist(obs$Freq, xlab = "# of Observations", main = "Histogram of # of Obs")
```

```{r}
# Score vs cultivar plot
boxplot(df.merge.bot$Score~df.merge.bot$Block,ylab = "Score",xlab = "Block")
boxplot(df.merge.bot$Score~df.merge.bot$HarvestPeriod,ylab = "Score",xlab = "Harvest")
ggplot(df.merge.bot, aes(x=HarvestPeriod, y=Score, fill=Block)) +
  geom_boxplot()+ theme_bw()

# number of observation in each block
# table(df.merge.bot$Block)
# table(df.merge.bot$HarvestPeriod)
```

```{r}
# Traits
hist(df.merge.bot$Brix, xlab = "Brix", main = "Histogram of Brix")
hist(df.merge.bot$CorrectedTitro, xlab = "Titro", main = "Histogram of Titro")
hist(df.merge.bot$Firmness.kg, xlab = "Firmness", main = "Histogram of Firmness")
```

## Linear mixed model with Gaussian assumption

$$Y_{ijk}=Cultivar_{ijk}+\beta_{1} Brix_{ijk}+\beta_{2}Titro_{ijk}+\beta_{3}Firmness_{ijk}+Harvest_{i}+Block_{ij}+\epsilon_{ijk}$$

, where $Y_{ijk}$ is the severity score of the disease for the $k^{th}$ strawberry in the $i^{th}$ harvest and $j^{th}$ block.$Cultivar_{ijk}, Brix_{ijk}, Titro_{ijk}, Firmness_{ijk}$ are the fixed effect. $Harvest_{i}$~$N(0,\sigma_{h})$ is a random effect. $Block_{ij}$~$N(0,\sigma_{b})$ is the random effect of the $j^{th}$ block nested in the $i^{th}$ harvest. $\epsilon_{ijk}$~$N(0,\sigma_{e})$ is the error term. The random effects and the error terms are independent.


```{r}
lmm1 = lmer(Score ~ Entry + Brix + CorrectedTitro + Firmness.kg + +(1 | HarvestPeriod)+(1 | HarvestPeriod:Block), data = df.merge.bot2)
AIC(lmm1)

par(mfrow=c(1,2))
plot(lmm1, type=c("p","smooth"), col.line=1, xlab = "Fitted Value",
     ylab = "Residual")
lattice::qqmath(lmm1)

s = summary(lmm1)
gt(as.data.frame(s[["coefficients"]]),rownames_to_stub = T)
```

## Mixed model with ordinal responses

`clmm` requires that all random effects to have at least 3 levels.



```{r}
fmm2 <- clmm(factor(Score) ~  Entry +Brix + CorrectedTitro + Firmness.kg +  (1 | HarvestPeriod:Block), data = df.merge.bot2, link = "logit") 
sf = summary(fmm2) 
coef = as.data.frame(sf[["coefficients"]])
coef[coef$`Pr(>|z|)`<=0.05,]
```

```{r}
drop1(fmm2, test = "Chi")
```

```{r}
fmm <- clmm(factor(Score) ~  Entry + Firmness.kg +  (1 | HarvestPeriod:Block), data = df.merge.bot2, link = "logit") 
summary(fmm)  
```

Interpretation:

We choose the lowest AIC or BIC as our best model, so logit link is used.

```{r}
link = c("logit", "probit", "cloglog", "loglog")
compare = data.frame(link = link,
                     AIC = rep(0,4),
                     BIC = rep(0,4))

for(i in 1:4){
    fmm <- clmm(factor(Score) ~  Entry +Brix + CorrectedTitro + Firmness.kg +  (1 | HarvestPeriod:Block), data = df.merge.bot2, link = link[i]) 
    compare[i,"AIC"] = AIC(fmm)
    compare[i,"BIC"] = BIC(fmm)
}
gt(compare)
```

Pro

Con: can only have 1 random effect

Discussion: 
- Few cultivars have significant p-values. We may consider increasing the sample size for each cultivar or combine some cultivars with similar genomes or characteristics into one group.
